---
title: "Multiple linear regression"
output: 
  learnr::tutorial:
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
knitr::opts_chunk$set(exercise.completion = FALSE)

knitr::opts_chunk$set(error = TRUE)

library(tidyverse)
library(openintro)
library(GGally)


```

## Grading the professor

Many college courses conclude by giving students the opportunity to evaluate 
the course and the instructor anonymously. However, the use of these student 
evaluations as an indicator of course quality and teaching effectiveness is 
often criticized because these measures may reflect the influence of 
non-teaching related characteristics, such as the physical appearance of the 
instructor. The article titled, "Beauty in the classroom: instructors' 
pulchritude and putative pedagogical productivity" by Hamermesh and Parker 
found that instructors who are viewed to be better looking receive higher 
instructional ratings. 

Here, you will analyse the data from this study in order to learn what goes 
into a positive professor evaluation.

## Getting Started

### Load packages

In this lab, you will explore and visualize the data using the **tidyverse** 
suite of packages. The data can be found in the companion package for OpenIntro resources, **openintro**.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(GGally)
```

This is the first time we're using the `GGally` package. You will be using the `ggpairs` function from this package later in the lab.

### The data

The data were gathered from end of semester student evaluations for a large 
sample of professors from the University of Texas at Austin. In addition, six 
students rated the professors' physical appearance. The result is a data 
frame where each row contains a different course and columns represent variables about the courses and professors. It's called `evals`.

```{r}
glimpse(evals)
```

We have observations on 21 different variables, some categorical and some 
numerical. The meaning of each variable can be found by bringing up the help file:

```{r help-evals}
?evals
```

#### Exercise
1.  Describe the distribution of `score`. Is the distribution skewed? What
    does that tell you about how students rate courses? Is this what you
    expected to see? Why, or why not?

```{r dist-score, exercise = TRUE}

```


```{r dist-score-solution}
ggplot(evals, aes(score)) + geom_histogram()

# the distribution is left-skewed, meaning we observe just a few low scores
# while the majority of observations are rather high score; 
# maybe not what you might have expected
```




2.  Excluding `score`, select two other variables and describe their
    relationship with each other using an appropriate visualisation.


```{r age-outfit, exercise = TRUE}

```


```{r age-outfit-solution}
# we take a look at the relation between age and pic_outfit; in particular,
# we analyse the age distribution with respect to pic_outfit

ggplot(evals, aes(x = pic_outfit, y = age)) + geom_boxplot()

# the average age seems to be lower in the group with non formal outfit
```



## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking teachers
are evaluated more favourably. Let's create a scatterplot to see if this appears 
to be the case:

```{r scatter-score-bty_avg}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_point()
```

Before you draw conclusions about the trend, compare the number of observations 
in the data frame with the approximate number of points on the scatterplot. 
Is anything awry?

#### Exercise
1. Replot the scatterplot, but this time use `geom_jitter` as your layer. What 
    was misleading about the initial scatterplot?


```{r scatter-score-bty_avg-jitter, exercise = TRUE}

```


```{r scatter-score-bty_avg-jitter-solution}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter()

# in the initial scatterplot a lot of overplotting was present; so, the number
# of observations was underestimated
```


2.  Let's see if the apparent trend in the plot is something more than
    natural variation. Fit a linear model called `m_bty` to predict average
    professor score by average beauty rating. Write out the equation for the linear 
    model and interpret the slope. Is the average beauty rating a statistically
    significant predictor? Does it appear to be a practically significant predictor?
    
Add the line of the bet fit model to your plot using the following:

```{r scatter-score-bty_avg-line-se, exercise=TRUE}


```
    
```{r scatter-score-bty_avg-line-se-solution}
m_bty <- lm(score ~ bty_avg, data = evals)
summary(m_bty)

# the model says that a unit increase in the average beauty rating leads
# on average to a score increase of roughtly 0.07
# the average beauty rating, is a significant predictor 

ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter() +
  geom_smooth(method = "lm")

#The blue line is the model. The shaded gray area around the line tells
# you about the variability you might expect in your predictions. To turn
# that off, use `se = FALSE`.

ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter() +
  geom_smooth(method = "lm", se = FALSE)

```


3.  Use residual plots to evaluate whether the conditions of least
 squares regression are reasonable. Provide plots and comments for each
 one (see the Simple Regression Lab for a reminder of how to make these).

```{r m_bty-residuals, exercise = TRUE}

```


```{r m_bty-residuals-solution}
plot(m_bty)

# the first plot shows no structre (which it shouldn't), the second plot
# shows slight skewness in the distribution of the residuals (which it
# shouldn't), the third plot shows points spreading homogeneously
# spreading around 1 (which should be the case), the fourth plots shows
# no outlier 

```



## Multiple linear regression

The data set contains several variables on the beauty score of the professor: 
individual ratings from each of the six students who were asked to score the 
physical appearance of the professors and the average of these six scores. Let's 
take a look at the relationship between one of these scores and the average 
beauty score.

```{r bty-rel}
ggplot(data = evals, aes(x = bty_f1lower, y = bty_avg)) +
  geom_point()

evals %>% 
  summarise(cor(bty_avg, bty_f1lower))
```

As expected, the relationship is quite strong---after all, the average score is 
calculated using the individual scores. You can actually look at the 
relationships between all beauty variables (columns 13 through 19) using the 
following command:

```{r bty-rels}
evals %>%
  select(contains("bty")) %>%
  ggpairs()
```

These variables are collinear (correlated), and adding more than one of these 
variables to the model would not add much value to the model. In this 
application and with these highly-correlated predictors, it is reasonable to use
the average beauty score as the single representative of these variables.

In order to see if beauty is still a significant predictor of professor score 
after you've accounted for the professor's gender, you can add the gender term 
into the model.

```{r scatter-score-bty_avg_pic-color}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
summary(m_bty_gen)
```


#### Exercise
1.  P-values and parameter estimates should only be trusted if the
    conditions for the regression are reasonable. Verify that the conditions
    for this model are reasonable using diagnostic plots. Try to use the function `autoplot()`. Remember that `autoplot()` relies on first running `library(ggfortify)`

```{r m_bty_gen-residuals, exercise = TRUE}

```


```{r m_bty_gen-residuals-solution}
library(ggfortify)
autoplot(m_bty_gen)

# the plots look similar as for the model m_bty; due to the skweness one might 
# have some doubts

```

2.  Is `bty_avg` still a significant predictor of `score`? Has the addition
    of `gender` to the model changed the parameter estimate for `bty_avg`?


```{r m_bty_gen-slope_bty_avg, exercise = TRUE}

```


```{r m_bty_gen-slope_bty_avg-solution}
# bty_avg is still a significan predictor of score

coef(m_bty)
coef(m_bty_gen)

# the estimate for bty_avg slightly increased compared to the model m_bty

```



Note that the estimate for `gender` is now called `gendermale`. You'll see this 
name change whenever you introduce a categorical variable. The reason is that R 
recodes `gender` from having the values of `male` and `female` to being an 
indicator variable called `gendermale` that takes a value of $0$ for 
female professors and a value of $1$ for male professors. (Such variables 
are often referred to as "dummy" variables.)

As a result, for female professors, the parameter estimate is multiplied by zero,
leaving the intercept and slope form familiar from simple regression.

\[
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \cdot bty\_avg + \hat{\beta}_2 \cdot 0 \\
&= \hat{\beta}_0 + \hat{\beta}_1 \cdot bty\_avg\end{aligned}
\]
The decision to call the indicator variable `gendermale` instead of 
`genderfemale` has no deeper meaning. R simply codes the category that 
comes first alphabetically as a $0$. (You can change the reference level of a categorical 
variable, which is the level that is coded as a 0, using the`relevel()` function. 
Use `?relevel` to learn more.)


3.  What is the equation of the line corresponding to the male professors? 
    For two professors who received the same beauty rating, which gender tends to
    have the higher course evaluation score?


```{r m_bty_gen-slope_gender, exercise = TRUE}

```


```{r m_bty_gen-slope_gender-solution}
# score= beta0 + beta1 * bty_avg + beta2 

# from
coef(m_bty_gen)
# we see that on average male professors have a score increase by roughly 0.17
# compared to female professors with the same average beauty rating
```




4. Create a new model called `m_bty_rank` with `gender` removed and `rank` 
    added in. Take a look at the estimated coefficients by using `coef()`. How does R appear to handle categorical variables that have more 
    than two levels? Note that the rank variable has three levels: `teaching`, 
    `tenure track`, `tenured`.
    
```{r m_bty_rank, exercise = TRUE}

```


```{r m_bty_rank-solution}
m_bty_rank <- lm(score ~ bty_avg + rank, data = evals)
coef(m_bty_rank)
# the category teaching is chosen as reference category; the output then shows
# two estimates for rank, reflecting the differences of the other two categories
# to the reference category
```



The interpretation of the coefficients in multiple regression is slightly 
different from that of simple regression. The estimate for `bty_avg` reflects
how much higher a group of professors is expected to score if they have a beauty
rating that is one point higher *while holding all other variables constant*. In
this case, that translates into considering only professors of the same rank 
with `bty_avg` scores that are one point apart.



## The search for the best model

We will start with a full model that predicts professor score based on rank, 
gender, ethnicity, language of the university where they got their degree, age, 
proportion of students that filled out evaluations, class size, course level, 
number of professors, number of credits, average beauty rating, outfit, and 
picture colour.


#### Exercise
Which variable would you expect to have the highest p-value in this model? 
    Why? *Hint:* Think about which variable would you expect to not have any 
    association with the professor score.

Let's run the model...

```{r m_full-model, tidy = FALSE}
m_full <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
```


#### Exercise
1. Check your suspicions from the previous exercise. Include the model output
    in your response.

```{r m_full, exercise = TRUE}

```


```{r m_full-solution}
summary(m_full)

# the highest p-value is observed for the number of professors teaching sections;
# you may decide on your own if this makes sense for you; the score should not
# be influenced by the fact if it is just one professor or a team teaching the course
```



2. Interpret the coefficient associated with the ethnicity variable.


```{r m_full-ethn, exercise = TRUE}

```


```{r m_full-ethn-solution}
coef(m_full)[5]

# while holding all other variables constant, professors who do not belong to
# an ethnic minority get on average higher scores by roughly 0.12 compared to
# professors from an ethnic minority

```



3. Drop the variable with the highest p-value and re-fit the model. Did the
    coefficients and significance of the other explanatory variables change?
    (One of the things that makes multiple regression interesting is that
    coefficient estimates depend on the other variables that are included in
    the model.) If not, what does this say about whether or not the dropped
    variable was collinear with the other explanatory variables?



```{r m_red, exercise = TRUE}

```


```{r m_red-solution}
m_red <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)

summary(m_red)

# the results for the reduced model are almost identical to the corresponding values
# in the full model; hence, there was almost no collinearity between cls_profs and
# the other variables
```


4. Using backward-selection and the AIC value as the selection criterion,
    determine the best model. 
    
```{r m_step, exercise = TRUE}

```


```{r m_step-solution}
m_step <- step(m_full, direction = "backward")
summary(m_step)

```

5. Verify that the conditions for this model are reasonable using diagnostic 
    plots.
    
```{r m_step-residuals, exercise = TRUE}

```


```{r m_step-residuals-solution}
plot(m_step)

# one still observes a slightly skewed distribution of the residuals, but everything
# else seems to be fine
```   
    


6. The original paper describes how these data were gathered by taking a
    sample of professors from the University of Texas at Austin and including 
    all courses that they have taught. Considering that each row represents a 
    course, could this new information have an impact on any of the conditions 
    of linear regression?
    
```{r letter-a, echo=FALSE}
question("Considering that each row represents a 
    course, could this new information have an impact on any of the conditions 
    of linear regression?",
  answer("No", message = "observations are assumed to be independent, which might not be the case"),
  answer("Yes", message = "observations are assumed to be independent, which might not be the case",correct = TRUE)
)
```



Based on your final model, we found the following characteristics of a professor and the course, which might lead on average to an increased score (at the University of Texas):

- The professor should be a male professor, should not be from an ethnic minority, should have an education from an English speaking school, should not be that old, should have a high beauty rating, should pick his outfit not to formal an should not provide a colour picture for determining his beauty rating

- The course evaluation should be completed by a high percentage of students and
the students should get just one credit for the course.


## References

* * *



<div id="license">
This tutorial is released under a 
[Creative Commons Attribution-ShareAlike 4.0 International License](http://creativecommons.org/licenses/by-sa/4.0/). 
This lab was adapted from an OpenIntro R Lab by Andrew Bray and Mine &Ccedil;etinkaya-Rundel.
</div>
