---
title: "Foundations for statistical inference - confidence intervals"
output: 
  learnr::tutorial:
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
knitr::opts_chunk$set(exercise.completion = FALSE)

knitr::opts_chunk$set(error = TRUE)

download.file("http://www.openintro.org/stat/data/ames.RData", destfile = "ames.RData")
load("ames.RData")

```



## Sampling from Ames, Iowa

If you have access to data on an entire population, say the size of every 
house in Ames, Iowa, it's straight forward to answer questions like, "How big 
is the typical house in Ames?" and "How much variation is there in sizes of 
houses?". If you have access to only a sample of the population, as is often 
the case, the task becomes more complicated. What is your best guess for the 
typical size if you only know the sizes of several dozen houses? This sort of 
situation requires that you use your sample to make inference on what your 
population looks like.

## The data

In the previous lab, "Sampling Distributions", we looked at the population data
of houses from Ames, Iowa. Let's start by loading that data set.

```{r load-data}
download.file("http://www.openintro.org/stat/data/ames.RData", destfile = "ames.RData")
load("ames.RData")
```

In this lab we'll start with a simple random sample of size 60 from the 
population. Note that the data set has information on many housing variables, but for the first portion of the lab we'll focus on the size of the house, represented by the variable `Gr.Liv.Area`.

```{r sample, echo=-1}
set.seed(19053)
samp <- data.frame(area = sample(ames$Gr.Liv.Area, 60))
```


#### Exercise

Describe the distribution of your sample. What would you say is the 
"typical" size within your sample? Also state precisely what you interpreted 
"typical" to mean.

```{r samp-dist-setup}
set.seed(19053)
samp <- data.frame(area = sample(ames$Gr.Liv.Area, 60))
```


```{r samp-dist, exercise = TRUE}


```


```{r samp-dist-solution}
summary(samp$area)
ggplot(samp, aes(x = area)) + geom_histogram(bins = 7)
# the distribution of the sample looks symmetric
# the typical size is around 1440 square feet
# a measure for "typical value" would be the sample mean for a symmetric
# distribution
```







## Confidence intervals

One of the most common ways to describe the typical or central value of a 
distribution is to use the mean. In this case we can calculate the mean of the 
sample using,

```{r sample-mean}
(sample_mean <- mean(samp$area))
```

Return for a moment to the question that first motivated this lab: based on 
this sample, what can we infer about the population? Based only on this single 
sample, the best estimate of the average living area of houses sold in Ames 
would be the sample mean, usually denoted as $\overline{x}$ (here we're calling it 
`sample_mean`). That serves as a good *point estimate* but it would be useful 
to also communicate how uncertain we are of that estimate. This can be 
captured by using a *confidence interval*.

We can calculate a 95% confidence interval for a sample mean by adding and 
subtracting 1.96 standard errors to the point estimate (See Section 4.2.3 if 
you are unfamiliar with this formula).

```{r ci}
se <- sd(samp$area) / sqrt(60)
lower <- sample_mean - 1.96 * se
upper <- sample_mean + 1.96 * se
c(lower, upper)
```

This is an important inference that we've just made: even though we don't know 
what the full population looks like, we're 95% confident that the true 
average size of houses in Ames lies between the values `r lower` and `r upper`. 
There are a few conditions that must be met for this interval to be valid.


#### Exercise

For the confidence interval to be valid, the sample mean must be normally 
distributed and have standard error $s / \sqrt{n}$. 



```{r shape-pop-dist, echo=FALSE}
question("What conditions must be met for this to be true?",
  answer("The sample observations are independent.", correct = TRUE),
  answer("The sample size is large: greater than 30 is a good rule of thumb.", correct = TRUE),
  answer("The sample size is at most 50"),
  answer("The population distribution is not strongly skewed.", correct = TRUE),
  answer("The sample distribution has to be left-skewed")
)
```




## Confidence levels

In this case we have the luxury of knowing the true population mean since we 
have data on the entire population. This value can be calculated using the 
following command:

```{r pop-mean}
mean(ames$Gr.Liv.Area)
```


#### Exercise

Does your confidence interval capture the true average size of houses in 
Ames? Use `<=` and `&` to check if the population mean is covered by you confidence interval.


```{r ci-popmean-setup}
set.seed(19053)
samp <- data.frame(area = sample(ames$Gr.Liv.Area, 60))
se <- sd(samp$area) / sqrt(60)
lower <- sample_mean - 1.96 * se
upper <- sample_mean + 1.96 * se
```


```{r ci-popmean, exercise = TRUE}


```


```{r ci-popmean-solution}
lower <= mean(ames$Gr.Liv.Area) & mean(ames$Gr.Liv.Area <= upper)
```








Using R, we're going to recreate many samples to learn more about how sample 
means and confidence intervals vary from one sample to another. The `infer` package comes in handy here.

Here is the rough outline:

-   Obtain 50 random samples.
-   Calculate for each sample the confidence interval.



The 50 samples are created with the `re_sample_n()` function from the `infer` package. In the next step we apply the `group_by()` function from the `dplyr` package. After that, each step is done seperately for each of the 50 samples. With
`summarise()` we can then compute the lower and upper bounds.



```{r resample, message=FALSE, warning=FALSE}
library(infer)
library(tidyverse)
n <- 60
ci_50 <- ames %>%
  select(Gr.Liv.Area) %>%
  rep_sample_n(size = 60, reps = 50) %>%
  group_by(replicate) %>%
  summarise(lower = mean(Gr.Liv.Area) - 1.96 * sd(Gr.Liv.Area) / sqrt(n),
            upper = mean(Gr.Liv.Area) + 1.96 * sd(Gr.Liv.Area) / sqrt(n),
            )
```


Lower bounds of these 50 confidence intervals are stored in the variable `lower`, 
and the upper bounds are in  the variable `upper`. Let's view the first interval.

```{r first-interval}
c(ci_50$lower[1], ci_50$upper[1])
```

* * *

## On your own

-   Using the following function (which was downloaded with the data set), 
    plot all intervals. What proportion of your confidence intervals include 
    the true population mean? Is this proportion exactly equal to the 
    confidence level? If not, explain why.

    ```{r plot-ci, eval=FALSE}
    plot_ci(ci_50$lower, ci_50$upper, mean(ames$Gr.Liv.Area))
    ```

-   Pick a confidence level of your choosing, provided it is not 95%. Calculate 50 confidence intervals at the chosen confidence level. You do not need to obtain new samples, simply calculate 
    new intervals based on the sample means and standard deviations you have 
    already collected. Using the `plot_ci` function, plot all intervals and 
    calculate the proportion of intervals that include the true population 
    mean. How does this percentage compare to the confidence level selected for
    the intervals?



<div id="license">
This tutorial is released under a 
[Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). 
This lab was adapted from an OpenIntro R Lab by Andrew Bray and Mine &Ccedil;etinkaya-Rundel.
</div>
