---
title: "Inference for numerical data"
output: 
  learnr::tutorial:
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
knitr::opts_chunk$set(exercise.completion = FALSE)

knitr::opts_chunk$set(error = TRUE)

download.file("http://www.openintro.org/stat/data/nc.RData", destfile = "nc.RData")
load("nc.RData")


```



## North Carolina births

In 2004, the state of North Carolina released a large data set containing 
information on births recorded in this state. This data set is useful to 
researchers studying the relation between habits and practices of expectant 
mothers and the birth of their children. We will work with a random sample of 
observations from this data set.

## Exploratory analysis

Load the `nc` data set into our workspace.

```{r load-data, eval=FALSE}
download.file("http://www.openintro.org/stat/data/nc.RData", destfile = "nc.RData")
load("nc.RData")
```

We have observations on 13 different variables, some categorical and some 
numerical. The meaning of each variable is as follows.

variable         | description
---------------- | -----------
`fage`           | father's age in years.
`mage`           | mother's age in years.
`mature`         | maturity status of mother.
`weeks`          | length of pregnancy in weeks.
`premie`         | whether the birth was classified as premature (premie) or full-term.
`visits`         | number of hospital visits during pregnancy.
`marital`        | whether mother is `married` or `not married` at birth.
`gained`         | weight gained by mother during pregnancy in pounds.
`weight`         | weight of the baby at birth in pounds.
`lowbirthweight` | whether baby was classified as low birthweight (`low`) or not (`not low`).
`gender`         | gender of the baby, `female` or `male`.
`habit`          | status of the mother as a `nonsmoker` or a `smoker`.
`whitemom`       | whether mom is `white` or `not white`.

#### Exercise 


```{r shape-pop-dist, echo=FALSE}
question("What are the cases in this data set? ",
  answer("fathers"),
  answer("babies", correct = TRUE),
  answer("mothers")
)
```


Determine the number of cases in `nc`

```{r nc-cases, exercise = TRUE}

```


```{r nc-cases-solution}
nrow(nc)
```



As a first step in the analysis, we should consider summaries of the data. This
can be done using the `summary` command:

```{r summary}
summary(nc)
```

As you review the variable summaries, consider which variables are categorical 
and which are numerical. For numerical variables, are there outliers? If you 
aren't sure or want to take a closer look at the data, make a graph.

Consider the possible relationship between a mother's smoking habit and the 
weight of her baby. Plotting the data is a useful first step because it helps 
us quickly visualize trends, identify strong associations, and develop research
questions.

#### Exercise 

Make a side-by-side boxplot of `habit` and `weight`. What does the plot 
highlight about the relationship between these two variables?


```{r mweight, exercise = TRUE}

```


```{r mweight-solution}
ggplot(nc, aes(x = habit, y = weight)) + geom_boxplot()
```



The box plots show how the medians of the two distributions compare, but we can
also compare the means of the distributions using the `group_by()` function from
`dplyr` to split the `weight` variable into the `habit` groups and then summarising the contained information by using the `mean()` function.

```{r by-means, warning=FALSE}
nc %>%
  group_by(habit) %>%
  summarise(mean_weight = mean(weight))
```

There is an observed difference, but is this difference statistically 
significant? In order to answer this question we will conduct a hypothesis test.

*Remark:* In the last example we also used the pipe operator `%>%`, which forwards the data from one step to the next one. 


## Inference

#### Exercise  

Check the sample sizes within each group. Remember that you can compute the group size using the summarise function `n()`. In addition check the shape of the distribution of `weight` for each group


```{r weight-group, exercise = TRUE}

```


```{r weight-group-solution}
nc %>%
  group_by(habit) %>%
  summarise(n())

# remove NA value
nc <- filter(nc, !is.na(habit))
ggplot(nc, aes(sample = weight)) + stat_qq() + stat_qq_line() +
  facet_wrap(~ habit)
```


Both distributions are (moderately) skewed. But due to the large sample size, we 
can still apply the testing framework of Chapter 5.




#### Exercise

Write the hypotheses for testing if the average weights of babies born to 
smoking and non-smoking mothers are different.


```{r hypoth-smoke, echo=FALSE}
question("The hypotheses for testing if the average weights of babies born to 
smoking and non-smoking mothers are different are:",
  answer("$H_0: \\mu_{\\text{non smoking mums}}\\not= \\mu_{\\text{smoking mums}}\\quad H_A: \\mu_{\\text{non smoking mums}} = \\mu_{\\text{smoking mums}}$"),
  answer("$H_0: \\mu_{\\text{non smoking mums}}= \\mu_{\\text{smoking mums}}\\quad H_A: \\mu_{\\text{non smoking mums}} < \\mu_{\\text{smoking mums}}$"),
  answer("$H_0: \\mu_{\\text{non smoking mums}}\\not= \\mu_{\\text{smoking mums}}\\quad H_A: \\mu_{\\text{non smoking mums}} > \\mu_{\\text{smoking mums}}$"),
  answer("$H_0: \\mu_{\\text{non smoking mums}}= \\mu_{\\text{smoking mums}}\\quad H_A: \\mu_{\\text{non smoking mums}} \\not= \\mu_{\\text{smoking mums}}$", correct = TRUE)
)
```


To test the above given hypothesis, we can use the `t.test()` function.


```{r inf-weight-habit-ht}
t.test(weight ~ habit, alternative = "two.sided", mu = 0, data = nc)
```

Let's pause for a moment to go through the arguments of this custom function. 
The first argument uses the formula notation `weight ~ habit`, which  specifies that we are interested in differences in mean `weight` due to `habit`. 
The second argument `alternative` can be `"less"`, `"greater"`, or `"two.sided"`. 
In our case we have two sided test, which is actually the default case. So, we could 
have skipped this argument. This is also true for `mu`, since 0 is the default value.
The last argument specifies the data object, which has to contain (in our case) the
variables `weight` and `habit`.



#### Exercise

Use the function `t.test()` to compute a two sided 99% confidence interval for the 
difference  $\mu_{\text{non smoking mums}} - \mu_{\text{smoking mums}}$ - take a look at `?t.test` to figure out how to adjust the confidence level. Can you
reject the null hypothesis at the significance level 0.01?



```{r weight-ci, exercise = TRUE}

```



```{r weight-ci-solution}
t.test(weight ~ habit, conf.level = 0.99, data = nc)$conf.int

# the null can't be rejected, since the interval covers 0
```


* * *

## On your own

-   Calculate a 95% confidence interval for the average length of pregnancies 
(`weeks`) and interpret it in context. Note that since you're doing inference 
on a single population parameter, you can't use the formula notation in `t.test()`.


-   Conduct a hypothesis test evaluating whether the average weight gained by 
younger mothers is different than the average weight gained by mature mothers.

-   Now, a non-inference task: Determine the age cutoff for younger and mature 
mothers. Use a method of your choice, and explain how your method works.




<div id="license">
This tutorial is released under a 
[Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). 
This lab was adapted from an OpenIntro R Lab by Andrew Bray and Mine &Ccedil;etinkaya-Rundel.
</div>
