---
title: "Foundations for statistical inference - Sampling distributions"
output: 
  learnr::tutorial:
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
knitr::opts_chunk$set(exercise.completion = FALSE)

knitr::opts_chunk$set(error = TRUE)

download.file("http://www.openintro.org/stat/data/ames.RData", destfile = "ames.RData")
load("ames.RData")

library(tidyverse)
```
---

In this lab, we investigate the ways in which the statistics from a random 
sample of data can serve as point estimates for population parameters.  We're 
interested in formulating a *sampling distribution* of our estimate in order 
to learn about the properties of the estimate, such as its distribution.

## The data

We consider real estate data from the city of Ames, Iowa.  The details of 
every real estate transaction in Ames is recorded by the City Assessor's 
office.  Our particular focus for this lab will be all residential home sales 
in Ames between 2006 and 2010.  This collection represents our population of 
interest.  In this lab we would like to learn about these home sales by taking 
smaller samples from the full population.  Let's load the data.

```{r load-data}
download.file("http://www.openintro.org/stat/data/ames.RData", destfile = "ames.RData")
load("ames.RData")
```

We see that there are quite a few variables in the data set, enough to do a 
very in-depth analysis.  For this lab, we'll restrict our attention to just 
two of the variables: the above ground living area of the house in square feet 
(`Gr.Liv.Area`) and the sale price (`SalePrice`).  


Let's look at the distribution of area in our population of home sales by 
calculating a few summary statistics and making a histogram.

```{r area, message=FALSE, warning=FALSE}
library(tidyverse)
summary(ames$Gr.Liv.Area)
ggplot(ames, aes(x = Gr.Liv.Area)) + geom_histogram()

```


#### Exercise

Describe this population distribution.
```{r shape-pop-dist, echo=FALSE}
question("The shape of the population distribution is",
  answer("left-skewed"),
  answer("symmetric"),
  answer("right-skewed", correct = TRUE)
)
```


## The unknown sampling distribution

In this lab we have access to the entire population, but this is rarely the 
case in real life.  Gathering information on an entire population is often 
extremely costly or impossible.  Because of this, we often take a sample of 
the population and use that to understand the properties of the population.

If we were interested in estimating the mean living area in Ames based on a 
sample, we can use the following command to survey the population.

```{r samp1}
samp1 <- data.frame(area = sample(ames$Gr.Liv.Area, 50))
```

This command collects a simple random sample of size 50 from the vector 
`ames$Gr.Liv.Area`, which is assigned to `samp1`.  This is like going into the City 
Assessor's database and pulling up the files on 50 random home sales.  Working 
with these 50 files would be considerably simpler than working with all 2930 
home sales.

#### Exercise 

Describe the distribution of this sample. How does it compare to the 
    distribution of the population?


```{r samp1-setup}
set.seed(190528)
samp1 <- data.frame(area = sample(ames$Gr.Liv.Area, 50))
```


```{r samp1-dist, exercise = TRUE}


```


```{r samp1-dist-solution}
summary(samp1$area)
ggplot(samp1, aes(x = area)) + geom_histogram(bins = 6)
# the distribution of the sample is still right-skewed but not as skewed as 
# the population distribution
```



If we're interested in estimating the average living area in homes in Ames 
using the sample, our best single guess is the sample mean.

```{r mean-samp1}
mean(samp1$area)
```

Depending on which 50 homes you selected, your estimate could be a bit above 
or a bit below the true population mean of 1499.69 square feet.  In general, 
though, the sample mean turns out to be a pretty good estimate of the average 
living area, and we were able to get it by sampling less than 3% of the 
population.

#### Exercise

 Take a second sample, also of size 50, and call it `samp2`.  How does the 
    mean of `samp2` compare with the mean of `samp1`?  Suppose we took two 
    more samples, one of size 100 and one of size 1000. Which would you think 
    would provide a more accurate estimate of the population mean?



```{r samp2-setup}
set.seed(190528)
samp1 <- data.frame(area = sample(ames$Gr.Liv.Area, 50))
```


```{r samp2, exercise = TRUE}

```


```{r samp2-solution}
samp2 <- data.frame(area = sample(ames$Gr.Liv.Area, 50))
mean(samp2$area) == mean(samp1$area)

# the sample of length 1000 would provide a more accurate estimate
```


Not surprisingly, every time we take another random sample, we get a different 
sample mean.  It's useful to get a sense of just how much variability we 
should expect when estimating the population mean this way. The distribution 
of sample means, called the *sampling distribution*, can help us understand 
this variability. In this lab, because we have access to the population, we 
can build up the sampling distribution for the sample mean by repeating the 
above steps many times. Here we will generate 5000 samples and compute the 
sample mean of each. This will be done using the function `rep_sample_n()` from
the `infer` package.

```{r resample}
library(infer)
sample_means_50 <- ames %>%
  select(Gr.Liv.Area) %>%
  rep_sample_n(size = 50, reps = 5000) %>%
  group_by(replicate) %>%
  summarise(avg_area = mean(Gr.Liv.Area))

ggplot(sample_means_50, aes(x = avg_area)) + geom_histogram()
```

If you would like to adjust the bin width of your histogram to show a little 
more detail, you can do so by changing the `bins` argument in `geom_histogram()`.

```{r hist-bins}
ggplot(sample_means_50, aes(x = avg_area)) + geom_histogram(bins = 50)
```



#### Exercise

Describe the sampling distribution, and be sure to specifically note its center.  Would you expect the distribution to change if we instead collected 50,000 sample 
means?



## Sample size and the sampling distribution

The sampling distribution that we computed tells us much about estimating 
the average living area in homes in Ames.  Because the sample mean is an 
unbiased estimator, the sampling distribution is centered at the true average 
living area of the the population, and the spread of the distribution 
indicates how much variability is induced by sampling only 50 home sales.

To get a sense of the effect that sample size has on our distribution, let's 
build up two more sampling distributions: one based on a sample size of 10 and 
another based on a sample size of 100.

```{r samp-10-100}
sample_means_10 <- ames %>%
  select(Gr.Liv.Area) %>%
  rep_sample_n(size = 10, reps = 5000) %>%
  group_by(replicate) %>%
  summarise(avg_area = mean(Gr.Liv.Area))

sample_means_100 <- ames %>%
  select(Gr.Liv.Area) %>%
  rep_sample_n(size = 100, reps = 5000) %>%
  group_by(replicate) %>%
  summarise(avg_area = mean(Gr.Liv.Area))

```

To see the effect that different sample sizes have on the sampling 
distribution, plot the three distributions on top of one another.

```{r plot-samps, tidy = FALSE, fig.show='hold', fig.align='center', warning=FALSE}
xlimits <- range(sample_means_10$avg_area)

ggplot(sample_means_10, aes(x = avg_area)) + geom_histogram(bins = 50) +
  xlim(xlimits)
ggplot(sample_means_50, aes(x = avg_area)) + geom_histogram(bins = 50) +
  xlim(xlimits)
ggplot(sample_means_100, aes(x = avg_area)) + geom_histogram(bins = 50) +
  xlim(xlimits)

```


With the `xlim()` function we specify the range of the x-axis of the histogram, and by setting it equal to `xlimits` for each histogram, we ensure that all three 
histograms will be plotted with the same limits on the x-axis.

#### Exercise

When the sample size is larger, what happens to the center?  What about the spread?



* * *
## On your own

So far, we have only focused on estimating the mean living area in homes in 
Ames.  Now you'll try to estimate the mean home price.

-   Take a random sample of size 50 from `price`. Using this sample, what is 
    your best point estimate of the population mean?

-   Since you have access to the population, simulate the sampling 
    distribution for $\overline{x}_{price}$ by taking 5000 samples from the 
    population of size 50 and computing 5000 sample means.  Store these means 
    in a vector called `sample_means_50`. Plot the data, then describe the 
    shape of this sampling distribution. Based on this sampling distribution, 
    what would you guess the mean home price of the population to be? Finally, 
    calculate and report the population mean.

-   Change your sample size from 50 to 150, then compute the sampling 
    distribution using the same method as above, and store these means in a 
    new vector called `sample_means_150`. Describe the shape of this sampling 
    distribution, and compare it to the sampling distribution for a sample 
    size of 50.  Based on this sampling distribution, what would you guess to 
    be the mean sale price of homes in Ames?

-   Of the sampling distributions from 2 and 3, which has a smaller spread?  If
    we're concerned with making estimates that are more often close to the 
    true value, would we prefer a distribution with a large or small spread?




<div id="license">
This tutorial is released under a 
[Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). 
This lab was adapted from an OpenIntro R Lab by Andrew Bray and Mine &Ccedil;etinkaya-Rundel.
</div>
